---
title: "BDA - Hierarchical model"
author: "Anonymous"
output: 
  pdf_document: 
    toc: yes
    toc_depth: 1
    keep_tex: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```

```{r, echo=FALSE}

# INCLUDE LIBRARIES

library(aaltobda)
library (mvtnorm)
library("rstan")
library("loo")
library("rstanarm")
library("bayesplot")

options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
Sys.setenv(LOCAL_CPPFLAGS = '-march=native')
```

```{r, echo=FALSE}
# CREATION OF THE DATASET: rows=years, columns=states; the number of fires are the sum of the fires of the months for each year
amazon=as.data.frame(amazon)
years=levels(factor(sort(amazon$year)))
states=levels(factor(sort(amazon$state)))
dataset=as.data.frame(matrix(rep(0,length(years)*length(states)), nrow=length(years), byrow=TRUE))
colnames(dataset)=states
rownames(dataset)=years
for (j in 1:length(states)) {
  for (i in 1:length(years)) {
    dataset[i,j] = sum(amazon[which(amazon$year==years[i] & amazon$state==states[j]),4])
  }
}
```



```{r, echo=TRUE}

# CREATION OF THE DATASET (x=groups, y=values), in order to apply the models
x = c(rep(1,20), rep(2,20), rep(3,20), rep(4,20),rep(5,20),rep(6,20),rep(7,20),rep(8,20),rep(9,20),rep(10,20),rep(11,20),
      rep(12,20),rep(13,20),rep(14,20),rep(15,20),rep(16,20),rep(17,20),rep(18,20),rep(19,20),rep(20,20),rep(21,20),rep(22,20),
      rep(23,20))
y=dataset[,1]
for (i in 2:23) {
  y=cbind(y,dataset[,i])
}
y=as.vector(y)

```


## Point 1

#### SEPARATE MODEL
$$ y_{ji} \sim N(\mu_j,\sigma_j)$$
```{r, echo=TRUE, tidy=FALSE}

# DEFINITION OF THE SEPARATE MODEL IN STAN

separate_code = "

data {
  int<lower=0> N;             // number of data points
  int<lower=0> K;             // number of groups
  int<lower=1,upper=K> x[N];  // group indicator
  vector[N] y;
}

parameters {
  vector[K] mu;               // group means
  vector<lower=0>[K] sigma;   //group stds
}

model {
  y ~ normal(mu[x], sigma[x]);
}

generated quantities {
  vector[K] y_state;
  vector[N] log_lik;                                

  for (i in 1:N)                                           
    log_lik[i] = normal_lpdf(y[i] | mu[x[i]], sigma[x[i]]);     
  
  for (i in 1:K)
    y_state[i]=normal_rng(mu[i], sigma[i]);
}

"

# DEFINITION OF THE DATA
data_s = list(
  N = 23*20,
  K = 23,
  x = x,
  y = y
)

# FIT OF THE MODEL IN STAN
fit_separate <- stan(
  model_code = separate_code,  # Stan program
  data = data_s,    # named list of data
  chains = 4,     # number of Markov chains
  iter =4000,    # total number of iterations per chain
  warmup=3000,
  cores = 2       # number of cores (could use one per chain)
)

samples_s = extract(object=fit_separate, permuted = TRUE, inc_warmup = FALSE, include = TRUE)

#dim(samples_s$log_lik)

```






#### POOLED MODEL
$$ y_i \sim N(\mu,\sigma)$$

```{r, echo=TRUE, tidy=FALSE}

# DEFINITION OF THE POOLED MODEL IN STAN

pooled_code = "

data {
  int<lower=0> N;      // number of data points
  vector[N] y;         //
}

parameters {
  real mu;             // common mean
  real<lower=0> sigma; // common std
}

model {
  y ~ normal(mu, sigma);
}


generated quantities { 
  real ypred;
  vector[N] log_lik;
  ypred = normal_rng(mu,sigma);
  for (i in 1:N)
    log_lik[i] = normal_lpdf(y[i] | mu, sigma);     // added line
}
"

# DEFINITION OF THE DATA
data_pooled = list(
  N = 23*20,
  y = y
)

# FIT OF THE MODEL IN STAN
fit_pooled <- stan(
  model_code = pooled_code,  # Stan program
  data = data_pooled,    # named list of data
  chains = 4,     # number of Markov chains
  iter =4000,    # total number of iterations per chain
  warmup=3000,
  cores = 2       # number of cores (could use one per chain)
)

samples_p = extract(object=fit_pooled, permuted = TRUE, inc_warmup = FALSE, include = TRUE)

#dim(samples_p$log_lik)

```







#### HIERARCHICAL MODEL
$$ y_{ji} \sim N(\overline{\mu}+\mu_j,\sigma)$$


```{r, echo=TRUE, tidy=FALSE}

# DEFINITION OF HIERARCHICAL MODEL IN STAN - NORMAL DISTRIBUTION OF FIRES

hierarchical_code = "

data {
  int<lower=0> N;           // number of data points
  int<lower=0> K;           // number of groups
  int<lower=1,upper=K> x[N]; // group indicator
  vector[N] y;              
}

parameters {
  real mu0;                 // prior mean
  real<lower=0> sigma0;     // prior std
  vector[K] mu;             // group means
  real<lower=0> sigma;      // common std
}

model {
  
  // using a normal distribution
  mu0 ~ normal(1500,1000);      // weakly informative prior //Poisson and negative binomial

  // https://datascienceplus.com/bayesian-regression-with-stan-beyond-normality/
  // https://mc-stan.org/docs/2_20/functions-reference/nbalt.html
 
  sigma0 ~ cauchy(0,4);     // weakly informative prior
  mu ~ normal(mu0, sigma0); // population prior with unknown parameters
  sigma ~ cauchy(0,4);       // weakly informative prior
  y ~ normal(mu[x], sigma);
}

generated quantities {
  real ypred;
  real mupred;
  vector[K] y_state;
  vector[N] log_lik; 

  mupred = normal_rng(mu0,sigma0);
  ypred = normal_rng(mupred, sigma);
  
  for (i in 1:N) 
    log_lik[i] = normal_lpdf(y[i] | mu[x[i]], sigma); 

  for (i in 1:K)
    y_state[i]=normal_rng(mu[i], sigma);

}
"
# DEFINITION OF THE DATA
data_hierarchical = list(
  N = 23*20,
  K = 23,
  x = x,
  y = y
)

# FIT OF THE MODEL IN STAN
fit_hierarchical <- stan(
  model_code = hierarchical_code,  # Stan program
  data = data_hierarchical,    # named list of data
  chains = 4,     # number of Markov chains
  iter =4000,    # total number of iterations per chain
  warmup=3000,
  cores = 2       # number of cores (could use one per chain)
)

samples_h = extract(object=fit_hierarchical, permuted = TRUE, inc_warmup = FALSE, include = TRUE)

#dim(samples_h$log_lik)

```


```{r, echo=TRUE, tidy=FALSE}

# DEFINITION OF HIERARCHICAL MODEL IN STAN - Poisson distribution of fires

## the values are not integer!! Check the . in the original dataset
#https://github.com/avehtari/BDA_R_demos/blob/master/demos_rstan/ppc/poisson-simple.stan

hierarchical_code_2 = "

data {
  int<lower=0> N;           // number of data points
  int<lower=0> K;           // number of groups
  int<lower=1,upper=K> x[N]; // group indicator
  int<lower=0> y[N];              
}

parameters {
  real<lower=0> lambda;     
}

model {
  lambda ~ exponential(0.2); //change this parameter
  y ~ poisson(lambda);
}

generated quantities {
  int<lower=0> y_rep[K]; 
  vector[N] log_lik; 
  
  for (i in 1:N) 
    log_lik[i] = poisson_lpmf(y[i] | lambda);

  for (i in 1:K) 
    y_rep[i] = poisson_rng(lambda); 
}
"
# DEFINITION OF THE DATA
data_hierarchical = list(
  N = 23*20,
  K = 23,
  x = x,
  y = y
)


# FIT OF THE MODEL IN STAN
fit_hierarchical_2 <- stan(
  model_code = hierarchical_code_2,  # Stan program
  data = data_hierarchical,    # named list of data
  chains = 4,     # number of Markov chains
  iter =4000,    # total number of iterations per chain
  warmup=3000,
  cores = 2       # number of cores (could use one per chain)
)

samples_h_2 = extract(object=fit_hierarchical_2, permuted = TRUE, inc_warmup = FALSE, include = TRUE)


```


```{r, echo=TRUE, tidy=FALSE}

# DEFINITION OF HIERARCHICAL MODEL IN STAN - Negative binomial distribution

## the values are not integer!! Check the . in the original dataset

hierarchical_code_3 = "

data {
  int<lower=0> N;           // number of data points
  int<lower=0> K;           // number of groups
  int<lower=1,upper=K> x[N]; // group indicator
  int<lower=0> y[N];              
}

parameters {
  real alpha; 
  real beta;
}

model {
  alpha ~ exponential(0.2); //change this
  beta ~ exponential(0.2); //change this
  y ~ neg_binomial(alpha, beta);
}

generated quantities {
  int<lower=0> y_rep[K]; 
  vector[N] log_lik; 
  
  for (i in 1:N) 
    log_lik[i] = neg_binomial_lpmf(y[i] | alpha, beta);

  for (i in 1:K) 
    y_rep[i] = neg_binomial_rng(alpha, beta); 
}
"
# DEFINITION OF THE DATA
data_hierarchical = list(
  N = 23*20,
  K = 23,
  x = x,
  y = y
)


# FIT OF THE MODEL IN STAN
fit_hierarchical_3 <- stan(
  model_code = hierarchical_code_3,  # Stan program
  data = data_hierarchical,    # named list of data
  chains = 4,     # number of Markov chains
  iter =4000,    # total number of iterations per chain
  warmup=3000,
  cores = 2       # number of cores (could use one per chain)
)

samples_h_3 = extract(object=fit_hierarchical_3, permuted = TRUE, inc_warmup = FALSE, include = TRUE)


```






### Point 2
Compute the PSIS-LOO elpd values and the k-values for each of the three models.


```{r, echo=TRUE, tidy=FALSE}

# SEPARATE MODEL
loo_separate=loo(fit_separate)
plot(loo_separate)

log_lik_s <- extract_log_lik(fit_separate, merge_chains = FALSE)
r_eff_s <- relative_eff(exp(log_lik_s))
loo_s <- loo(log_lik_s, r_eff = r_eff_s, save_psis=TRUE, cores=2 )
print(loo_s)


# POOLED MODEL
loo_pooled=loo(fit_pooled) 
plot(loo_pooled)

log_lik_p <- extract_log_lik(fit_pooled, merge_chains = FALSE)
r_eff_p <- relative_eff(exp(log_lik_p))
loo_p <- loo(log_lik_p, r_eff = r_eff_p, save_psis=TRUE, cores=2 )
print(loo_p)


# HIERARCHICAL MODEL - normal
loo_hierarchical=loo(fit_hierarchical)
plot(loo_hierarchical)

log_lik_h <- extract_log_lik(fit_hierarchical, merge_chains = FALSE)
r_eff_h <- relative_eff(exp(log_lik_h))
loo_h <- loo(log_lik_h, r_eff = r_eff_h, save_psis=TRUE, cores=2 )
print(loo_h)


# HIERARCHICAL MODEL - poisson
loo_hierarchical_2=loo(fit_hierarchical_2)
plot(loo_hierarchical_2)

log_lik_h_2 <- extract_log_lik(fit_hierarchical_2, merge_chains = FALSE)
r_eff_h_2 <- relative_eff(exp(log_lik_h_2))
loo_h_2 <- loo(log_lik_h_2, r_eff = r_eff_h_2, save_psis=TRUE, cores=2 )
print(loo_h_2)


# HIERARCHICAL MODEL - negative binomial
loo_hierarchical_3=loo(fit_hierarchical_3)
plot(loo_hierarchical_3)

log_lik_h_3 <- extract_log_lik(fit_hierarchical_3, merge_chains = FALSE)
r_eff_h_3 <- relative_eff(exp(log_lik_h_3))
loo_h_3 <- loo(log_lik_h_3, r_eff = r_eff_h_3, save_psis=TRUE, cores=2 )
print(loo_h_3)


```


```{r, echo=TRUE, tidy=FALSE}

# ESTIMATION OF PSIS-LOO values 

loo_s$estimates[1] #separate
loo_p$estimates[1] #pooled
loo_h$estimates[1] #hierarchical - normal
loo_h_2$estimates[1] #hierarchical - poisson
loo_h_3$estimates[1] #hierarchical - negative binomial

```



### Point 3
Compute the effective number of parameters Peff for each of the three models.
https://mc-stan.org/loo/reference/loo-glossary.html

Answer:
Equation (7.15) in the book:
$$ p_{loo-cv} = lppd - lppd_{loo-cv} $$
where

$$ lppd_{loo-cv} $$ is the PSIS-LOO value (sum of the LOO log densities) (calculated above in point 2)

and 
$$ lppd = \sum_{i=1}^n \log( \frac{1}{S} \sum_{s=1}^S p(y_i|\theta^s))$$
The lppd of the observed data y is an overestimate of the elppd for future data.



```{r, echo=TRUE, tidy=FALSE}
S=4000
n=20*23

# ESTIMATION OF lppd

# SEPARATE
vector_s=rep(0,n)
for(i in 1:n)
 vector_s[i]=log(1/S*(sum(exp(samples_s$log_lik[,i]))))

# POOLED
vector_p=rep(0,n)
for(i in 1:n)
 vector_p[i]=log(1/S*(sum(exp(samples_p$log_lik[,i]))))


# HIERARCHICAL - normal
vector_h=rep(0,n)
for(i in 1:n)
 vector_h[i]=log(1/S*(sum(exp(samples_h$log_lik[,i]))))

# HIERARCHICAL - poisson
vector_h_2=rep(0,n)
for(i in 1:n)
 vector_h_2[i]=log(1/S*(sum(exp(samples_h_2$log_lik[,i]))))

# HIERARCHICAL - negative binomial
vector_h_3=rep(0,n)
for(i in 1:n)
 vector_h_3[i]=log(1/S*(sum(exp(samples_h_3$log_lik[,i]))))


```


```{r, echo=TRUE, tidy=FALSE}

#RESULTING VALUES FOR peff

peff_s = sum(vector_s) - loo_s$estimates[1]
peff_p = sum(vector_p) -loo_p$estimates[1]
peff_h = sum(vector_h) - loo_h$estimates[1]
peff_h_2 = sum(vector_h_2) - loo_h_2$estimates[1]
peff_h_3 = sum(vector_h_3) - loo_h_3$estimates[1]

peff_s
peff_p
peff_h #hierarchical normal
peff_h_2 #hierarchical poisson
peff_h_3 #hierarchical negative binomial

```






